Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2015-11-20 pldi-2016/paper.tex
fix 3

\noindent
This shows that provided types are correct with respect to the preferred shape relation. 
This is an obligation upon the author of a type provider, which is why
type soundness is ``relative''.
Our key theorem states that, for any input (which is a subshape of any of the samples) and 
any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
Using standard syntactic type safety  \cite{syntactic}, we prove type preservation 
(reduction does not change type) and progress (an expression can be reduced).

@@

The F\# Data library has been widely adopted by the industry and is one of the most downloaded
F\# libraries.\footnote{At the time of writing, the library has over 80,000 downloads on NuGet 
(package repository), 1,821 commits and 44 contributors on GitHub.} A practical demonstration of 
development using the library can be seen in an attached screencast and additional documentation can be
found at \url{http://fsharp.github.io/FSharp.Data}.

@@

The \emph{relativized safety} property does not guarantee the same amount of safety as traditional
closed-world type safety, but it reflects the reality of programming with external data that is 
increasingly important \cite{age-of-web}. Type providers do not reduce the safety -- they simply 
illuminate the existing issues.

One such issue is the handling of schema change. With type providers, the sample is captured 
at compile-time. If the schema changes later (so that the input is no longer related to the  
shape of the sample), the program fails at runtime and developers have to handle the exception. 
The same problem happens when using weakly-typed techniques.

F\# Data can help discover such errors earlier. Our first example (\S\ref{sec:introduction})
points the JSON type provider at a sample using a live URL. This has the advantage that a re-compilation 
fails when the schema changes, which is an indication that the program needs to be updated to reflect the
change.

In general, there is no better solution for plain XML, CSV and JSON data sources. However, some data 
sources provide versioning support (with meta-data about how the schema changed). For those, a type 
provider could adapt automatically, but we leave this for future work.

@@

\paragraph{Reading XML documents.}
Mapping XML documents to structural values is perhaps the most interesting. For each node, we
create a record. Attributes become record fields and the body becomes a field with a special
name:
%
{\small{
\begin{verbatim}
  <root id="1">
    <item>Hello!</item>
  </root>    
\end{verbatim}
}}
%
\noindent
This XML becomes a record \ident{root} with fields \ident{id} and $\bullet$ for the body. 
The nested element contains only the $\bullet$ field with the inner text. As with CSV, we
infer shape of primitive values:
%
\begin{equation*}
\ident{root}~\{ \ident{id} \mapsto 1, \bullet \mapsto [ \ident{item}~\{ \bullet \mapsto \str{Hello!} \}] \}
\end{equation*}
%
When generating F\# types for XML documents, we also lifts the members nested under the $\bullet$ 
field into the parent type to simplify the structure of the generated type.

@@

When introducing type providers (\S\ref{sec:providers-sum}), we mentioned that F\# Data 
implements a special handling of heterogeneous collections. This allows us to avoid inferring a 
labelled top shapes in many common scenarios. In the earlier example, a sample collection 
contains a record (with \strf{pages} and \strf{page} fields) and a nested collection with values.

@@

C$\omega$ \cite{comega-xs} is the most similar to F\# Data. It extends C\# with types similar to our
structural types (including similar heterogeneous collections), but it does not infer the types
from a sample. It also extends the type system of the host language (rather than using a general purpose
embedding).

\paragraph{Advanced type systems.}
A number of other advanced type system features could be used to tackle the problem discussed 
in this paper. The Ur \cite{ur} language has a rich system for working with records; 
meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming \cite{multi-stage}
could be used to generate code for the provided types; and gradual typing \cite{gradual,gradual-js} 
can add typing to existing dynamic languages. As far as we are aware, none of these 
systems have been used to provide the same level of integration with XML, CSV and JSON.

\paragraph{Typing real-world data.}
Recent work \cite{typing-json} infers a succinct type of large JSON datasets using MapReduce.
It fuses similar types based on similarity. This is more sophisticated than our technique, but it 
makes formal specification of safety (Theorem~\ref{thm:safety}) difficult. Extending our 
\emph{relativized safety} to \emph{probabilistic safety} is an interesting future work.

The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
\cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
many of the patterns we explore in this paper, but formally specifying the safety property would be
more challenging.

@@

We explored the F\# Data type providers for XML, CSV and JSON. As most real-world data do not come 
with an explicit schema, the library uses \emph{shape inference} that deduces a shape from a set of 
samples. Our inference algorithm is based on a preferred shape relation. It prefers records to 
encompass the open world assumption and support developer tooling. The algorithm is predictable, which is 
important as developers need to understand how changing the samples affects the resulting types.

We explore the theory behind type providers. F\# Data is a prime example of 
type providers, but our work demonstrates a more general point. The types generated by type 
providers can depend on external input and so we can only guarantee \emph{relativized safety}, 
which says that a program is safe only if the actual inputs satisfy additional conditions.

Type providers have been described before, but this paper is novel in that it explores the 
properties of type providers that represent the ``types from data'' approach. Our experience suggests 
that this significantly broadens the applicability of statically typed languages to real-world 
problems that are often solved by weakly-typed techniques.

>>>>

\noindent
This shows that provided types are correct with respect to the preferred shape relation. 
In general, this is an obligation upon the author of a type provider, which is a second reason why
the type soundness is ``relative''.
Our key theorem states that, for any input (which is a subshape of any of the samples) and 
any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
Using standard syntactic type safety  \cite{syntactic}, we prove type preservation 
(reduction does not change type) and progress (an expression can be reduced).

@@

The F\# Data library has been widely adopted by users and is one of the most downloaded
F\# libraries.\footnote{At the time of writing, the library has over 80,000 downloads on NuGet 
(package repository), 1,821 commits and 44 contributors on GitHub.} A practical demonstration of 
development using the library can be seen in an attached screencast and additional documentation can be
found at \url{http://fsharp.github.io/FSharp.Data}.

@@

The \emph{relativized safety} property does not guarantee safety in the same way as traditional
closed-world type safety, but it reflects the reality that programming with external data is
increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of programming,
seen by example in \S\ref{sec:introduction}.

One such issue is the handling of sample change. With type providers, the sample is captured 
at compile-time. If the sample changes later (so that the input is no longer related to the  
shape of the sample), the program can fail at runtime and developers have to handle the exception. 
The same problem happens when using weakly-typed code with explicit failure cases.

F\# Data can help discover such errors earlier. Our first example (\S\ref{sec:introduction})
points the JSON type provider at a sample using a live URL. This has the advantage that a re-compilation 
fails when the sample changes, which is an indication that the program needs to be updated to reflect the
change. Repeated type-checking of programs by continuous integration systems can be used to help 
detect such changes.

In general, schema-free XML, CSV and JSON data sources will necessarily require techniques akin
to those we have shown. However, some data sources provide schema and schema versioning support 
(with meta-data about how the schema changed). For those, a type 
provider could be written to adapt automatically, but we leave this for future work.

@@

\paragraph{Reading XML documents.}
Mapping XML documents to structural values is more interesting. For each node, we
create a record. Attributes become record fields and the body becomes a field with a special
name:
%
{\small{
\begin{verbatim}
  <root id="1">
    <item>Hello!</item>
  </root>    
\end{verbatim}
}}
%
\noindent
This XML becomes a record \ident{root} with fields \ident{id} and $\bullet$ for the body. 
The nested element contains only the $\bullet$ field with the inner text. As with CSV, we
infer shape of primitive values:
%
\begin{equation*}
\ident{root}~\{ \ident{id} \mapsto 1, \bullet \mapsto [ \ident{item}~\{ \bullet \mapsto \str{Hello!} \}] \}
\end{equation*}
%
When generating F\# types for XML documents, we also lifts the members nested under the $\bullet$ 
field into the parent type to simplify the structure of the generated type.

@@

When introducing type providers (\S\ref{sec:providers-sum}), we mentioned how F\# Data 
handles heterogeneous collections. This allows us to avoid inferring 
labelled top shapes in many common scenarios. In the earlier example, a sample collection 
contains a record (with \strf{pages} and \strf{page} fields) and a nested collection with values.

@@

C$\omega$ \cite{comega-xs} bears similarity to F\# Data. Like many other
languages, it extends a host language with advanced types similar to our
structural shapes (including similar heterogeneous collections), but it 
does not infer the types from a sample. In contrast, in F\# Data 
we have avoided extending F\# at all, and the simplicity of the Foo 
target language shows we have avoided placing strong requirements 
(such as structural records) on the host language. Instead we provide nominal 
types based on structure, rather than adding an advanced
system of structural types into the host language.

\paragraph{Advanced type systems with meta-programming.}
A number of other advanced type system features could be used to tackle the problem discussed 
in this paper. The Ur \cite{ur} language has a rich system for working with records; 
meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming \cite{multi-stage}
could be used to generate code for the provided types; and gradual typing \cite{gradual,gradual-js} 
can add typing to existing dynamic languages. As far as we are aware, none of these 
systems have been used to provide the same level of integration with XML, CSV and JSON.

\paragraph{Typing real-world data.}
Recent work \cite{typing-json} infers a succinct type of large JSON datasets using MapReduce.
It fuses similar types based on similarity. This is more sophisticated than our technique, but it 
makes formal specification of safety (Theorem~\ref{thm:safety}) difficult. Extending our 
\emph{relativized safety} to \emph{probabilistic safety} is an interesting future direction.

The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
The schema definitions in PADS are similar to our shapes. The structure inference for LearnPADS
\cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
many of the patterns we explore in this paper, but formally specifying the safety property would be
more challenging.

@@

We explored the F\# Data type providers for XML, CSV and JSON. As most real-world data does not come 
with an explicit schema, the library uses \emph{shape inference} that deduces a shape from a set of 
samples. Our inference algorithm is based on a preferred shape relation. It prefers records to 
encompass the open world assumption and support developer tooling. The algorithm is predictable, which is 
important as developers need to understand how changing the samples affects the resulting types.

We explored the theory behind type providers. F\# Data is a prime example of 
type providers, but our work demonstrates a more general point. The types generated by type 
providers can depend on external input and so we can only guarantee \emph{relativized safety}, 
which says that a program is safe only if the actual inputs satisfy additional conditions.

Type providers have been described before, but this paper is novel in that it explores the 
properties of type providers that represent the ``types from data'' approach. Our experience suggests 
that this significantly broadens the applicability of statically typed languages to real-world 
problems that are often solved by error-prone weakly-typed techniques.
